Investigating the Gaussianity of Supernova SALT2 Summary Statistics

### 2016-08-15 dessn/investigations/gaussianity/paper/gaussianity.tex
Some more updates

One key assumption when doing this is that these summary statistics are in fact gaussian in nature. Often this is explicitly assumed in the analysis \citep{MarchTrotta2011, RubinAldering2015}, however it is often implicitly assumed \citep{SullivanGuy2011, SuzukiRubin2012, CampbellDAndrea2013, BetouleKessler2014}. Investigation into asymmetries in the summary statistics are generally limited to asymmetries in the parent colour and stretch distribution \citep{ScolnicRest2014}. {\red More citations.} In our paper we therefore do not investigate biases or asymmetries found in underlying population distributions, but asymmetries and biases in the SALT2 parameter probability surface, and the impact of assuming the Gaussianity of said surface.

@@

We investigate the potential asymmetries in SALT2 summary statistics by fitting the light curves and obtaining summary statistics using different algorithms in the software package \verb|sncosmo| \citep{Barbary2014}. Specifically, we obtain a full posterior surface using our own MCMC fitting methodology, and compare against summary statistics generated using a different MCMC approach (the \verb|mcmc_lc| method in \verb|sncosmo|), nestled sampling (the \verb|nest_lc| method) and minuit (via the \verb|fit_lc| method). The MCMC based algorithm utilises the package \verb|emcee| \citep{ForemanMackey2013}, whilst the nestled sampling is based off \verb|nestle| \citep{Skilling2004, Barbary2015}, and both algorithms should map out the full posterior surface, giving summary statistics from the mean and covariance of that full surface. The third algorithm, invoked via \verb|fit_lc| uses the minuit algorithm \citep{James1975, minuit2015}, and as such is a maximum likelihood estimator and does not map out the full posterior surface. For a Gaussian surface, all three algorithms should agree, however for non-Gaussian surfaces the minuit algorithm will be expected to produce different output. As such, we test the SALT2 probability surfaces for skewness and fit with the four algorithms described previously. The summary statistics are converted into distance modulus via the Philips relation
\begin{align}
\mu = m_B - M + \alpha x_1 - \beta c,
\end{align}
where we adopt the fiducial $\alpha = 0.14$ and $\beta = 3.15$ values from \citet{BetouleKessler2014}. This allows us to directly see the effects of any skewness or non-Gaussianity in any parameter on the cosmologically important distance modulus. We can compare the distributions in $\mu$ for the full posterior surface and for summary statistics, and also verify the skewness of the full distribution of $\mu$. We investigate any shift in $\langle \mu \rangle$, and change in the standard deviation $\sigma_\mu$, binning by redshift, with results illustrated in Figure~\ref{fig:bias_des}. We find strong evidence that, with increasing redshift, the skewness of the SALT2 fit posterior increases, and causes the maximum likelihood estimator to not only produce biased output, but overestimate the uncertainty of the result.
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias_dessky.pdf}
    \caption{The mean bias in {\red fill out caption when updated figure}}
    \label{fig:bias_des}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias_dessky_deep.pdf}
	\caption{The mean bias in {\red fill out caption when updated figure}}
	\label{fig:bias_des_deep}
\end{figure}
The likelihood surface for one of the skewed supernova fits is shown as an example in Figure \ref{fig:example}, where the effect of the skewness is readily apparent.
\begin{figure*}
	\includegraphics[width=\textwidth]{../output/bias_surface.png}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:example}
\end{figure*}
To confirm that the skewness is a result of redshift and not simply a result of the general trend of decreasing signal to noise, we plot the skewness as a function of both redshift and signal to noise in Figure \ref{fig:bias_skew}. Interestingly, doing this shows a negative skewness for low redshift, low signal-to-noise light curves, and a positive skewness for high redshift light curves, regardless of signal-to-noise.
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:bias_skew}
\end{figure}
Having confirmed that skewness becomes significant at higher redshift, we now seek to characterise this potential impact on cosmology.

\section{Cosmological Impact}
\label{sec:cosmology}
Fit cosmology against simulated SN

Do for des shallow and deep, using survey area to produce z dist plus extra low-z sample (0.05<z<0.2). simulate from light curves, apply selection effect of ston>5to cull bad fits. have <x> shallow.

toy cosmology model. adopt $\alpha$ and $\beta$ values for Phillips correction from Betoule2014, assume known gaussian intrinsic scatter, have $\Omega_m$, $w$ and $M_B$ as cosmology parameters - marginalise over $M_B$.

Show cosmology results.

\begin{figure}
	\includegraphics[width=\columnwidth]{../output/obs_cosmology_shallow.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:obs_cosmology_shallow}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/obs_cosmology_deep.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:obs_cosmology_deep}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/comparison_shallow.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:comparison_shallow}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/comparison_deep.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:comparison_deep}
\end{figure}

%\begin{figure}
%	% To include a figure from a file named example.*
%	% Allowable file formats are eps or ps if compiling using latex
%	% or pdf, png, jpg if compiling using pdflatex
%	\includegraphics[width=\columnwidth]{example}
%    \caption{This is an example figure. Captions appear below each figure.
%	Give enough detail for the reader to understand what they're looking at,
%	but leave detailed discussion to the main body of the text.}
%    \label{fig:example_figure}
%\end{figure}
%
%% Example table
%\begin{table}
%	\centering
%	\caption{This is an example table. Captions appear above each table.
%	Remember to define the quantities, symbols and units used.}
%	\label{tab:example_table}
%	\begin{tabular}{lccr} % four columns, alignment for each
%		\hline
%		A & B & C & D\\
%		\hline
%		1 & 2 & 3 & 4\\
%		2 & 4 & 6 & 8\\
%		3 & 5 & 7 & 9\\
%		\hline
%	\end{tabular}
%\end{table}

>>>>

One key assumption when doing this is that these summary statistics are in fact Gaussian in nature. Often this is explicitly assumed in the analysis \citep{MarchTrotta2011, RubinAldering2015}, however it is often implicitly assumed \citep{SullivanGuy2011, SuzukiRubin2012, CampbellDAndrea2013, BetouleKessler2014}. Investigation into asymmetries in the summary statistics are generally limited to asymmetries in the parent colour and stretch distribution \citep{ScolnicRest2014}. {\red More citations.} In our paper we therefore do not investigate biases or asymmetries found in underlying population distributions, but asymmetries and biases in the SALT2 parameter probability surface, and the impact of assuming the Gaussianity of said surface.

@@

We investigate the potential asymmetries in SALT2 summary statistics by fitting the light curves and obtaining summary statistics using different algorithms in the software package \verb|sncosmo| \citep{Barbary2014}. Specifically, we obtain a full posterior surface using our own MCMC fitting methodology, and compare against summary statistics generated using a different MCMC approach (the \verb|mcmc_lc| method in \verb|sncosmo|), nestled sampling (the \verb|nest_lc| method) and minuit (via the \verb|fit_lc| method). The MCMC based algorithm utilises the package \verb|emcee| \citep{ForemanMackey2013}, whilst the nestled sampling is based off \verb|nestle| \citep{Skilling2004, Barbary2015}, and both algorithms should map out the full posterior surface, giving summary statistics from the mean and covariance of that full surface. The third algorithm, invoked via \verb|fit_lc| uses the minuit algorithm \citep{James1975, minuit2015}, and as such is a maximum likelihood estimator and does not map out the full posterior surface. For a Gaussian surface, all three algorithms should agree, however for non-Gaussian surfaces the minuit algorithm will be expected to produce different output. As such, we test the SALT2 probability surfaces for skewness and fit with the four algorithms described previously. The summary statistics are converted into distance modulus via the Philips relation
\begin{align}
\mu = m_B - M + \alpha x_1 - \beta c,
\end{align}
where we adopt the fiducial $\alpha = 0.14$ and $\beta = 3.15$ values from \citet{BetouleKessler2014}. This allows us to directly see the effects of any skewness or non-Gaussianity in any parameter on the cosmologically important distance modulus. We can compare the distributions in $\mu$ for the full posterior surface and for summary statistics, and also verify the skewness of the full distribution of $\mu$. We investigate any shift in $\langle \mu \rangle$, and change in the standard deviation $\sigma_\mu$, binning by redshift, with results illustrated in Figure~\ref{fig:bias_des}. We find strong evidence that, with increasing redshift, the skewness of the SALT2 fit posterior increases, and causes the maximum likelihood estimator to not only produce biased output, but overestimate the uncertainty of the result.
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias_dessky.pdf}
    \caption{The bias in distance modulus $\mu$ for observations generated as per the DES shallow field. Supernova results for all algorithms were binned into redshift groups (top two subplots) and the full posterior surface was analysed for skewness (bottom subplot). We find evidence for skewness at higher redshifts, leading to diverging results between mean and maximum likelihood algorithms.}
    \label{fig:bias_des}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias_dessky_deep.pdf}
	\caption{The bias in distance modulus $\mu$ for observations generated as per the DES deep field. Supernova results for all algorithms were binned into redshift groups (top two subplots) and the full posterior surface was analysed for skewness (bottom subplot). We find evidence for skewness at higher redshifts, however the increased observational time (and subsequent increase in signal-to-noise) results in less biased results than the shallow field.}
	\label{fig:bias_des_deep}
\end{figure}
%The likelihood surface for one of the skewed supernova fits is shown as an example in Figure \ref{fig:example}, where the effect of the skewness is readily apparent.

%\begin{figure*}
%	\includegraphics[width=\textwidth]{../output/bias_surface.png}
%	\caption{The posterior surfaces for all algorithms in the SALT2 fit parameters $x_0$, $x_1$, $c$ and $t$, along with the derived value $\mu$. The skewness of the marginalised colour distribution has a significant effect on the skewness of the final $\mu$ distribution, which leads to disagreement between the algorithms used.}
%	\label{fig:example}
%\end{figure*}
To confirm that the skewness is a result of redshift and not simply a result of the general trend of decreasing signal to noise, we plot the skewness as a function of both redshift and signal to noise in Figure \ref{fig:bias_skew}. Interestingly, doing this shows a negative skewness for low redshift, low signal-to-noise light curves, and a positive skewness for high redshift light curves, regardless of signal-to-noise.
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/bias.pdf}
	\caption{The deviation between the maximum likelihood value for $\mu$ generated by minuit and the mean of the full posterior surfaces. Each point on the plot represents a simulated supernova, with the background surface showing a simple second order polynomial fit to the surface. We can see that, whilst the bias does decrease as signal-to-noise increases, it is still present for the high redsfhit supernova.}
	\label{fig:bias_skew}
\end{figure}
Having confirmed that skewness becomes significant at higher redshift, we now seek to characterise this potential impact on cosmology.

\section{Cosmological Impact}
\label{sec:cosmology}

\subsection{Uncorrected Biases}
For our initial cosmology comparison we generate a thousand supernova for both the shallow and deep fields using simulation cosmology of $\Omega_m = 0.30$, $w = -1.0$, $M_B = -19.3$ and $\sigma_{\rm int} = 0.1$. 10\% of the supernova are drawn from a linear redshift distribution ($0.05 < z < 0.2$) included emulate an external dataset used to constrain the low redshift range of the Hubble diagram, with the remaining 90\% of supernova drawn from a redshift distribution following DES survey volume and $z < 1.2$. The thousand supernova all have the selection effect of having at least one point above a signal-to-noise of 5. We take these thousand supernova and fit their light curves (generated with the same cadence, $t_0$ and weather effects as discussed in the previous section) using both MCMC methods and maximum likelihood methods. As the bias comes from a cut before fitting the light curves, both the set of summary statistics from MCMC and maximum likelihood fitting have the same bias, and any difference between fits to the different sets of summary statistics is thus a product of divergence between the two methodologies and not the cause of selection effects.  We fix $\alpha$ and $\beta$ as done previously, set $\sigma_{\rm int} = 0.1$, and fit a simple model of $\Omega_m$, $w$ and $M_B$.

\begin{figure}
	\includegraphics[width=\columnwidth]{../output/obs_cosmology_shallow.pdf}
	\caption{A Hubble diagram sowing the underlying simulation cosmology and the resulting 1000 supernova distance moduli for observations in the DES shallow field. Each supernova is shown twice, once when using summary statistics based in the mean to calculate the distance modulus, and then when using maximum likelihood to generate summary statistics and calculate the distance modulus. Selection effects heavily prune high redshift supernova.}
	\label{fig:obs_cosmology_shallow}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/obs_cosmology_deep.pdf}
	\caption{A Hubble diagram sowing the underlying simulation cosmology and the resulting 1000 supernova distance moduli for observations in the DES deep field. Each supernova is shown twice, once when using summary statistics based in the mean to calculate the distance modulus, and then when using maximum likelihood to generate summary statistics and calculate the distance modulus. The increased observation in the deep field gives us more high redshift supernova than the shallow field.}
	\label{fig:obs_cosmology_deep}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/comparison_shallow.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:comparison_shallow}
\end{figure}
\begin{figure}
	\includegraphics[width=\columnwidth]{../output/comparison_deep.pdf}
	\caption{{\red Fill out caption in updated figure}}
	\label{fig:comparison_deep}
\end{figure}

Figures~\ref{fig:obs_cosmology_shallow} and \ref{fig:obs_cosmology_deep} show the Hubble diagrams (and simulation cosmology) for the shallow and deep simulated supernova respectively. Visible in the diagrams is the systematic bias between the two methodologies at high redshift. Marginalising over the absolute magnitude $M_B$, the cosmological surfaces are presented in Figures~\ref{fig:comparison_shallow} and \ref{fig:comparison_deep}. The shallow field results show a highly significant bias, with even the deep field showing approximately a $1\sigma$ shift when using different methodologies for generating the summary statistics. Table \ref{tab:unbias_comp} shows the marginalised parameter summaries for comparison, and the shift between summary statistics is significant.

\begin{table}
	\centering
	\caption{Comparison between the shallow and deep field cosmology for the mean and maximum likelihood summary statistics.}
	\label{tab:unbias_comp}
    \begin{tabular}{cccc}
       	\hline
       	Model & $\Omega_m$ & $w$ & $M$ \\ 
       	\hline
       	Deep MCMC & $0.317^{+0.027}_{-0.038}$ & $-1.08^{+0.12}_{-0.15}$ & $-19.290^{+0.016}_{-0.017}$ \\ 
       	Deep Max. Like. & $0.362^{+0.024}_{-0.030}$ & $-1.29^{+0.17}_{-0.15}$ & $-19.299^{+0.014}_{-0.020}$ \\ 
       	Shallow MCMC & $0.341^{+0.035}_{-0.051}$ & $-1.17^{+0.19}_{-0.17}$ & $-19.289^{+0.014}_{-0.017}$ \\ 
       	Shallow Max. Like. & $0.426^{+0.018}_{-0.031}$ & $-1.59^{+0.23}_{-0.18}$ & $-19.306^{+0.014}_{-0.018}$ \\ 
       	\hline
    \end{tabular}
\end{table}

The current investigation is not sufficient grounds to claim that any previous cosmology analyses that use maximum likelihood statistics are biased, as this depends strongly on the method of bias correction utilised in prior analyses. For example, methods in which biases are computed from the underlying cosmology may be prone to biases, whilst methods rely on adjusting the data to take into account selection effects calibrated with simulations \citep{KesslerBecker2009, ConleyGuy2011} may be consistent, given simulation SALT2 parameter distributions are drawn from distributions also fit using the observed maximum likelihood summary statistics. Similarly, Bayesian methods, such as modelling selection effects as a Gaussian CDF \citep{RubinAldering2015} are algorithm independent and may correct both sets of summary statistics equally well.

\subsection{Simulation Bias Correction}

Try and do it the Rick Kessler way, verify everything is a-ok

\subsection{Modelled Corrections}

Try and do it the Rubin et al way

\subsection{Cosmological Corrections}

Try and do it the Alex Kim way.
